msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-13 paper.tex

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Two developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing technology---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2014scrm}
were necessary to simulate whole chromosomes.

@@

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2014scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

@@

\subsection*{Simulation interface}
The primary interface for \msprime\ is through a Python application
programming interface (API). This has many advantages. One is
that we can grow the interface over time. Another is that
downstream applications such as \texttt{stdpopsim}~\citep{adrion2020community}
can be developed.

The \texttt{msp simulate} command also supports demographic models
specified in Demes format~\citep{gower2021demes}, allowing users with
no Python experience to run complex demographic simulations.

@@

A significant advantage to simulating the coalescent with recombination
backwards in time using Hudson's algorithm rather than
left-to-right
approaches~\citep{wiuf1999recombination,chen2009fast,staab2014scrm}
is the ease with which additional processes can be integrated.
We have seen that processes such as gene conversion, complex
demography, instantaneous bottlenecks, multiple merger caolescents
and selective sweeps have been integrated seamlessly into
the simulations. Incorporating any of these extensions into
either the exact or approximate left-to-right processes would require
complex derivations. Because we are going backwards
in time, processes can be modelled as competing exponentials
using Gillespie's algorithm \jkcomment{I think?},
and therefore in principle we can
add as many extra processes as we like.
\jkcomment{so, backwards in time
sims in msprime is extensible for the future.}

>>>>

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Two developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing technology---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.

@@

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

@@

\subsection*{Simulation interface}
The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces such as \ms\ make it very easy to run simple
simulations, but as model complexity and the number of parameters specified increases
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from some distribution. In practice, for any significant simulation
project (or at least, any reproducible one), users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and packages
such \texttt{coala}~\citep{staab2016coala} have been developed
to allow simulations be run directly in a high-level
scripting language.

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}. This
approach greatly increases the flexibility of simulations, and allows all
processing for many simulation analyses to be performed in a single script. The
primary interface for \msprime\ is through a well documented and stable Python
API, which has allowed a number of downstream tools to be built on
% JK: must be more, right?
top of it~\citep{adrion2020community,rivera2021simulation}
as well as numerous easily reproducible simulation
% TODO: more papers with exemplorary reproducible simulation pipelines based
% on msprime!
pipelines~\citep[e.g.][]{martin2017human}.
We also provide an \ms\ compatible
command line interface to support existing workflows, as well as the
\texttt{msp simulate} command that supports specifying demographic models in
Demes format~\citep{gower2021demes}.

A major change for the 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
resolving a major source of confusion and error. The 0.x APIs are fully
supported and tested, and will be maintained for the foreseeable future.

@@

A significant advantage to simulating the coalescent with recombination
backwards in time using Hudson's algorithm rather than
left-to-right
approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
is the ease with which additional processes can be integrated.
We have seen that processes such as gene conversion, complex
demography, instantaneous bottlenecks, multiple merger caolescents
and selective sweeps have been integrated seamlessly into
the simulations. Incorporating any of these extensions into
either the exact or approximate left-to-right processes would require
complex derivations. Because we are going backwards
in time, processes can be modelled as competing exponentials
using Gillespie's algorithm \jkcomment{I think?},
and therefore in principle we can
add as many extra processes as we like.
\jkcomment{so, backwards in time
sims in msprime is extensible for the future.}
